
---
title: "BIOL_3100 Final Project"
subtitle: "Education & Career Success; Data Analysis and Prediction Model"
author: Myles Ashbaker
output:
  html_document:
    toc: true
    toc_float: true
    theme: yeti
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 5)
library(tidyverse)
library(janitor)
library(GGally)
library(caret)
library(easystats)
set.seed(123)
```

# Introduction

This analysis examines which academic and experiential factors predict early career success using the **Education & Career Success** dataset (*n = 5,000*).

Specifically:

1. **Starting Salary Prediction** – applying multiple linear regression to identify the key drivers of initial compensation (e.g., GPA, internships, soft skills).  
2. **Entrepreneurial Propensity** – using logistic regression and threshold tuning to understand which attributes influence the likelihood of pursuing entrepreneurship after graduation.

The goal of this study is to provide actionable insights for students and advisors on the most impactful factors for salary and entrepreneurial outcomes.

# Data Loading & Cleaning

### Thought Process
- Standardize column names for tidy handling  
- Remove the identifier column  
- Convert categorical variables to factors for proper modeling

```{r data-load}
df <- read_csv("../../Data/MRA_education_career_success.csv", show_col_types = FALSE)

df <- janitor::clean_names(df)

df$student_id <- NULL

df <- df %>%
  mutate(
    entrepreneurship   = factor(entrepreneurship, levels = c("No","Yes")),
    gender             = factor(gender),
    field_of_study     = factor(field_of_study),
    current_job_level  = factor(current_job_level)
  )

glimpse(df)
```

# Exploratory Data Analysis

### Pairwise Numeric Scatterplots
```{r eda-pairs, fig.width=12, fig.height=10}
num_cols <- names(df)[vapply(df, is.numeric, logical(1))]
GGally::ggpairs(df, columns = num_cols, aes(alpha = 0.4))
```

### Salary Distribution
##### shows most graduates earn in the mid‑range, with a long tail of higher salaries but very few low salaries.
```{r eda-salary-dist}
ggplot(df, aes(starting_salary)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  scale_x_continuous(labels = scales::label_dollar()) +
  labs(title = "Distribution of Starting Salary",
       x = "Starting Salary", y = "Number of Graduates") +
  theme_minimal()
```

### Salary by Field of Study
##### reveals that STEM majors cluster at higher pay levels, while arts‑related fields sit lower on the scale.
```{r eda-field}
ggplot(df, aes(fct_reorder(field_of_study, starting_salary, median), starting_salary)) +
  geom_boxplot(outlier.alpha = 0.3) +
  coord_flip() +
  scale_y_continuous(labels = scales::label_dollar()) +
  labs(title = "Starting Salary by Field of Study",
       x = NULL, y = "Starting Salary") +
  theme_minimal()
```

### Internships vs Salary
##### illustrates a clear upward trend: more internships generally translate into a higher starting salary.
```{r eda-internship}
ggplot(df, aes(internships_completed, starting_salary)) +
  geom_jitter(width = 0.25, alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE, color = "darkred") +
  scale_y_continuous(labels = scales::label_dollar()) +
  labs(title = "Internships Completed vs. Starting Salary",
       x = "Internships Completed", y = "Starting Salary") +
  theme_minimal()
```

### Years to Promotion by Entrepreneurship
##### indicates that entrepreneurs and non‑entrepreneurs reach their first promotion in a similar time frame.
```{r eda-promotion}
ggplot(df, aes(entrepreneurship, years_to_promotion, fill = entrepreneurship)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Years to Promotion by Entrepreneurship Status",
       x = "Entrepreneurship", y = "Years to Promotion") +
  theme_minimal()
```

### Salary vs University Ranking
##### displays a gentle downward slope showing graduates from better‑ranked universities (lower rank number) tend to earn more.
```{r eda-ranking}
ggplot(df, aes(university_ranking, starting_salary)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "loess", se = FALSE, color = "steelblue") +
  scale_y_continuous(labels = scales::label_dollar()) +
  labs(title = "Starting Salary vs. University Ranking",
       x = "University Ranking (lower = better)", y = "Starting Salary") +
  theme_minimal()
```

### Salary by Gender
##### box‑plots suggest median salaries are broadly comparable across genders with overlapping ranges.
```{r eda-gender}
ggplot(df, aes(gender, starting_salary, fill = gender)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  scale_y_continuous(labels = scales::label_dollar()) +
  labs(title = "Starting Salary by Gender",
       x = "Gender", y = "Starting Salary") +
  theme_minimal() +
  theme(legend.position = "none")
```

### Distribution of Job Offers
##### confirms that most graduates receive one to three offers, while very high counts are rare.
```{r eda-offers}
ggplot(df, aes(job_offers)) +
  geom_bar(fill = "coral", alpha = 0.8) +
  labs(title = "Distribution of Number of Job Offers",
       x = "Job Offers", y = "Number of Graduates") +
  theme_minimal()
```

### Career Satisfaction by Entrepreneurship
##### shows entrepreneurs report slightly higher median career satisfaction than non‑entrepreneurs.
```{r eda-satisfaction}
ggplot(df, aes(entrepreneurship, career_satisfaction, fill = entrepreneurship)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Career Satisfaction by Entrepreneurship Status",
       x = "Entrepreneurship", y = "Career Satisfaction (1–10)") +
  theme_minimal() +
  theme(legend.position = "none")
```

# Train/Test Split
##### Thought process: To measure how well the models generalize to new graduates, prevents over‑fitting, and gives an honest estimate of real‑world predictive performance.
```{r split}
train_idx <- createDataPartition(df$starting_salary, p = 0.8, list = FALSE)
train <- df[train_idx, ]
test  <- df[-train_idx, ]
```

# Modeling

### Linear Regression for Starting Salary

##### Thought Process: Start with a theory‑driven subset, compare to a full model, and let stepwise AIC select the most parsimonious predictor set.

```{r lm-fit}
lm_base <- lm(starting_salary ~ university_gpa + internships_completed +
                certifications + soft_skills_score + networking_score +
                field_of_study, data = train)

lm_full <- lm(starting_salary ~ . -entrepreneurship, data = train)

lm_step <- MASS::stepAIC(lm_full, trace = 0)  # AIC‑optimized model

compare_performance(lm_base, lm_step, rank = TRUE)
model_parameters(lm_step)

rmse_val <- RMSE(predict(lm_step, test), test$starting_salary)
cat("Test RMSE:", round(rmse_val, 2), "\n")
```

#### Diagnostics
```{r lm-diag, fig.height=6}
par(mfrow = c(2,2)); plot(lm_step); par(mfrow = c(1,1))
```

### Logistic Regression for Entrepreneurship

##### Thought Process:  Use a full logistic model, perform stepwise AIC, then tune the probability cutoff to balance sensitivity and specificity.

```{r glm-fit}
glm_full <- glm(entrepreneurship ~ . -starting_salary, data = train, family = binomial)
glm_step <- MASS::stepAIC(glm_full, trace = 0)

model_parameters(glm_step, exponentiate = TRUE)

probs <- predict(glm_step, test, type = "response")

thresholds <- seq(0.1, 0.9, by = 0.1)
metrics <- map_dfr(thresholds, function(t) {
  pred <- factor(ifelse(probs >= t, "Yes", "No"), levels = c("No","Yes"))
  cm <- confusionMatrix(pred, test$entrepreneurship, positive = "Yes")
  tibble(threshold = t,
         Accuracy = cm$overall["Accuracy"],
         Sensitivity = cm$byClass["Sensitivity"],
         Specificity = cm$byClass["Specificity"])
})
print(metrics)
```

```{r threshold-plot}
metrics %>%
  pivot_longer(-threshold) %>%
  ggplot(aes(threshold, value, color = name)) +
  geom_line() + geom_point() +
  labs(title = "Threshold Tuning Metrics",
       x = "Probability Threshold", y = "Metric Value") +
  theme_minimal()
```

```{r glm-final}
best_t <- metrics$threshold[which.max(metrics$Sensitivity + metrics$Specificity)]
cat("Optimal threshold:", best_t, "\n")

pred_final <- factor(ifelse(probs >= best_t, "Yes", "No"), levels = c("No","Yes"))
conf_final <- confusionMatrix(pred_final, test$entrepreneurship, positive = "Yes")
print(conf_final)
```

# Summary  

- **Data preparation:** Cleaned variable names (janitor::clean_names), removed the ID column, and converted categorical variables to factors, leaving 19 variables for 5,000 graduates.  
- **Exploratory analysis:** Visualised numeric relationships (pairwise plots) and key univariate distributions. Plots showed clear positive trends between internships, soft‑skills, and starting salary, and strong salary separation by field of study with STEM majors leading.  

- **Model workflow**
 - **Salary:** Compared a theory‑driven baseline model to a full model, then used stepwise AIC to obtain a parsimonious linear model. Test RMSE ≈ **`r scales::dollar(round(rmse_val, 0))`**.
  
  - **Entrepreneurship:** Fit a full logistic model, applied stepwise AIC, then tuned the probability threshold to balance performance (optimal ≈ `r best_t`). Achieved test accuracy ≈ `r round(metrics$Accuracy[metrics$threshold==best_t], 3)` 

# Conclusions  

**Salary Insights** 

- **Experiential learning** (internships) and **interpersonal abilities** (soft skills, networking) have the largest positive impact on starting salary. 
- Choice of major matters - **STEM graduates** consistently command higher pay.  
- The final model’s RMSE (about **`r scales::dollar(round(rmse_val, 0))`**) indicates reasonably accurate salary prediction across diverse backgrounds.  

**Entrepreneurship Insights** 

- **Academic strength** (higher high‑school GPA) and **personal outlook** (career satisfaction) modestly increase the odds of pursuing entrepreneurship.  
- Using a tuned cutoff (about `r best_t`) balances sensitivity and specificity, providing roughly **60 percent** correct identification for both entrepreneurs and non‑entrepreneurs.  

**Actionable Takeaways**  

- Students: Prioritise **internships** and soft‑skill development** to boost early‑career earnings; nurture **satisfaction and academic excellence** if entrepreneurship is a goal.  

- Advisors and Universities: Facilitate **practical experiences and communication training**, especially for non‑STEM majors, to narrow salary gaps and support entrepreneurial pathways.  



**These findings suggest students maximize early‑career salary by investing in experiential learning and interpersonal skills, while a strong academic record and satisfaction mindset correlate with entrepreneurial intent.**

